---
title: "IntervsMono"
output: html_document
date: "2025-07-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries}
library(ggplot2)
library(multcompView)
library(lmerTest)
library(lme4)
library("PerformanceAnalytics")
library(dplyr)
library(psych)
library(readxl) 
library(emmeans)
library(multcompView)
library(multcomp)
library(readr)
```

```{r}
data <- read_csv("NewYork/data/early_life_intermono.csv")

data$ENT <- as.factor(data$ENT)
data$Entry <- as.factor(data$Entry)
data$Block <- as.factor(data$Block)
data$Serpentine <- as.character(data$Serpentine)
data$treatment <- as.factor(data$treatment)
data$`Seed/Plot g` <- as.numeric(data$`Seed/Plot g`)
data$coated <- as.factor(data$coated)
data$Chalcids <- as.factor(data$Chalcids)
```

```{r}
# re-level the treatment for the models so that mono is the reference instead of intercrop
# Compared to monoculture baseline to test whether intercrop improves performance relative to monoculture
# check levels
#levels(data$treatment) # first level is the current reference
# re-level
#data$treatment <- relevel(data$treatment, ref = "monoculture")
```

##  Are values in treatment (inter vs mono) different

Main effects: 
Entry: differences between entries in intercrop, relative to the reference Entry (IWAF_H_MN1) in intercrop
treatment: change going from intercrop to monoculture 
interaction: do entries respond differently to monoculture than they do in intercrop
   
# Fall Stand Count
```{r Fstandcount}
mix.model <- lmer(Fall.Alfalfa.Stand.Count ~ Entry * treatment + (1 | Block), data = data)
anova(mix.model)
# Entry is significant Pr(>F) = 0.0001585, treatment marginally so Pr(>F) = 0.0534388, and the interaction is not significant Pr(>F) = 0.1511384
## stand count differed significantly by entry
## no strong evidence of an entry-by-treatment interaction across all entries; most entries respond similarly to treatment
summary(mix.model)
# Entries IWAF_H_WI1, IWAF_L_KS1, IWAF_L_MN1, IWAF_L_WI1, UMN5560, and UMN5561 were not different than the  reference entry (IWAF_H_MN1) in monoculture
# Larry15 was significantly lower stand count than the reference entry (IWAF_H_MN1) in monoculture Pr(>|t|)=0.00974
# No overall difference between intercrop and monoculture for the reference entry Pr(>|t|) =0.33807
# Entry IWAF_L_KS1, IWAF_L_MN1, Larry15, and UMN5560 were not different from the reference entry (IWAF_H_MN1)
## these entries responded the same in intercrop as in monoculture 
# Entry UMN5561 was marginally lower stand count
## may be negativly effected by intercrop
# Entries IWAF_H_WI1 and IWAF_L_WI1 were significanly lower stand counts in intercrop compared to monoculture 
```

# Fall Height 
```{r Fheight}
mix.model <- lmer(Fall.Alfalfa.Height ~ Entry * treatment + (1 | Block), data = data)
anova(mix.model)
# Entry is significant Pr(>F) = 0.001157, treatment and the interaction are not significant
## hieght differed significantly by entry
## no strong evidence of an entry-by-treatment interaction across all entries; most entries respond similarly to treatment
summary(mix.model)
# Larry15 was significantly lower height than the reference entry (IWAF_H_MN1) in monoculture Pr(>|t|)=0.00778
```

# Winter Survival
```{r winter_survival}
mix.model <- lmer(Alfalfa.winter.survival ~ Entry * treatment + (1 | Block), data = data)
anova(mix.model)
# Entry is marginally significant Pr(>F) = 0.06699, treatment and the interaction are not significant
## no strong evidence for significance of entry or treatment;  entries respond similarly to treatment
summary(mix.model)
# Entry UMN5561 had significanly higher winter survival than the reference entry (IWAF_H_MN1) in monoculture Pr(>|t|)=0.00778
# one individual interaction is significant : UMN5561:monoculture Pr(>|t|) = 0.03803
# this entry shows a significant treatment effect where winter survival was lower in intercrop compared with monoculture 
```

# Spring Stand Count
```{r Sstandcount}
mix.model <- lmer(Spring.Alfalfa.stand.count ~ Entry * treatment + (1 | Block), data = data)
anova(mix.model)
# Entry, treatment and the interaction are not significant
## no strong evidence for significant effect of entry or treatment or thier interaction;  entries respond similarly to treatment
summary(mix.model)
```

# Spring Vigor
```{r Svigor}
mix.model <- lmer(Spring.Alfalfa.vigor ~ Entry * treatment + (1 | Block), data = data)
anova(mix.model)
# Entry is significant Pr(>F) = 0.001463, treatment and the interaction are not significant
## vigor differed significantly by entry
## no strong evidence of an entry-by-treatment interaction across all entries; most entries respond similarly to treatment
summary(mix.model)
# Entry Larry had marginally lower vigor than the reference entry (IWAF_H_MN1) in monoculture Pr(>|t|)=0.05486
# Entries IWAF_H_WI1 and UMN5561 had significanly lower vigor than the reference entry (IWAF_H_MN1) in monoculture

# Entry IWAF_L_MN1 is marginally significant Pr(>|t|) = 0.0875; all else equal, this entry has a vigor 1.2500 units higher than the intercept (IWAF_H_MN1)
# two individual interaction terms are marginally significant : IWAF_H_WI1:monoculture Pr(>|t|) = 0.0620 and UMN5561:monoculture Pr(>|t|) = 0.0929
## a few specific entries show a significant treatment effect
## both entries show a weaker responce to monoculture compared to the reference
```

```{r}
mix.model <- lmer(Spring.Alfalfa.height ~ Entry * treatment + (1 | Block), data = mono_entry_data)
anova(mix.model)
# Entry, treatment and the interaction have no significant effect on spring alfalfa hieght
## no strong evidence for significance of entry or treatment;  entries respond similarly to treatment
summary(mix.model)
# interaction of IWAF_L_KS1:monoculture is marginally significant Pr(>|t|) = 0.0999; all else equal, this entry has a height 7.0500 units higher than the intercept (IWAF_H_MN1)
## Entry might respond positivly to monoculture treatment 
```
